Towards a Neural Co-Processor Which Restores Movement After Stroke: Modeling a Proof-of-Concept

### 2021-11-12 paper/article.tex
Training update
Paper progress

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.pdf}
	    \caption{Example muscle trajectories during a single trial, for one hand and shoulder muscle}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.pdf}
	    \caption{Proportion of loss attributable to hand muscles}
	\end{subfigure}
	\hfill
	\caption{Lesion designs which prevent forward propagation of object shape information
	         differentially impact hand pose. Loss here is an L2 distance
		 measured relative to the circuit's trajectories prior to the lesion.}
	\label{fig:lesion}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics[width=0.45\textwidth]{stim_single.png}
	\caption{Spatial and temporal smoothing of a single 16 dimensional,
	         randomized $\theta$, onto 100 simulated neurons.
		 Color values indicate the magnitude of the value summed into
		 each neuron's hidden state in that time step. After $t=0$,
		 $\theta$ is the zero vector.}
	\centering
	\label{fig:stim_single}
\end{figure}

@@

We achieve this through a simple error backpropagation, using PyTorch's
implementation of the $Adam$ optimizer. With each trial, task loss is
calculated and backpropagated into the mRNN, allowing it to adapt to
its inputs, one of which is the stimulation being applied. The learning
rate for the optimizer is set arbitrarily to a relatively low rate of
$1e-7$.

\subsection{Training Algorithm}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network.
The loss function when training the EN is the MSE loss between the actual and
predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN output, and target output, to the CPN. See Fig.
\ref{fig:training}.

TODO: explain EN training

@@

Alternate training EN and CPN.
\begin{itemize}
	\item Train an EN
		\begin{itemize}
			\item Training the EN until predictive power reaches a threshold, based on current task performance
			\item Training regime based on most recent CPN, similar CPNS (noise added to params), and white noise stimulation
			\item Single batch of data from applying current CPN set and white noise to the brain, EN fit to ``offline'' it over maining epochs
			\item Predictive power alone is not sufficient for use in training the CPN: we need to form the batch as above, otherwise training is unstable.
			\item White noise and noised params
			\item ** Learning instability
			\item ** Predictive power isn't enough
			\item ** Illustrate this
		\end{itemize}
	\item Train the CPN
		\begin{itemize}
			\item Backprop through the EN, effectively using the EN's prediction as ground truth
		\end{itemize}

	\item Data recycling/reuse
\end{itemize}

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{pseudocode.png}
\caption{Pseudocode of the Training Method}
\centering
\label{fig:pseudocode}
\end{figure}

@@

Using the simulation described above, we perform three experiments to
demonstrate the co-processor's ability to learn:
\begin{itemize}
	\item 50\% AIP loss
	\item 100\% connectivity lesion between the F5 and M1 modules
	\item 50\% M1 loss
\end{itemize}

@@

\section{Results}

>>>>

\begin{figure}[h]
	\centering
	\begin{subfigure}[c]{0.62\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_trajs.pdf}
	    \caption{Example muscle trajectories during a single trial, for one hand and shoulder muscle}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.32\textwidth}
	    \centering
	    \includegraphics[width=\textwidth]{lesion_hand.pdf}
	    \caption{Ratio of hand muscle MSE losses vs. overall MSE losses, for different lesion types}
	\end{subfigure}
	\hfill
	\caption{Lesion designs which prevent forward propagation of object shape information
	         differentially impact hand pose. Loss here is an L2 distance
		 measured relative to the circuit's trajectories prior to the lesion.}
	\label{fig:lesion}
\end{figure}

@@

\begin{figure}[h]
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single.png}
		\caption{$\sigma=1.875$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[c]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{stim_single2.png}
		\caption{$\sigma=2.875$}
	\end{subfigure}

	\caption{Spatial and temporal smoothing of a single 16 dimensional,
	         randomized $\theta$, onto 100 simulated neurons.
		 Color values indicate the magnitude of the value summed into
		 each neuron's hidden state in that time step. After $t=0$,
		 $\theta$ is the zero vector.}
	\label{fig:stim_single}
\end{figure}

@@

We achieve this through a simple error backpropagation, using PyTorch's
implementation of the $Adam$ optimizer. With each trial, task loss is
calculated and backpropagated into the mRNN, allowing it to adapt to
its inputs, one of which is the stimulation being applied. The learning
rate for the optimizer is set arbitrarily to a relatively low rate of
$1e-7$. The learning rate was not chosen in a principled way, since
we are not aware of a principled way of choosing it. However, we reason
that it should be lower than the co-processor by orders of magnitude due
to the rapidity by which an ANN can be retrained.

\subsubsection{Simulating recovery prior to co-processor use}
We sought to additonally pre-train the mRNN after applying the lesion in
order to simulate stroke recovery. For simulated lesions which zero the outputs
of neurons, the mRNN is, unfortunately, not an appropriate model.
We know that strokes can cause damage from which the subject, in many cases,
cannot recover. The mRNN model has sufficient redundancy built into it that
lesioning the outputs of neurons leaves enough remaining degrees of freedom
that a nearly full recovery can occur, unless so many neurons are lesioned
that no stimulation could be effective. We go into this further in the
Discussion section below.

We can, however, simulate pre-recovery in the case of a lesion that excludes
communication between the F5 and M1 modules. In that case, object shape information
cannot propagate forward in the network, in order to condition the hand for
grasping. The mRNN can, in pre-recovery, learn a stereotyped grasp, at which
point the co-processor's job will be to forward-propagate that information.
We explore this in one of our experiments, explained below.

\subsection{Training Algorithm}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network's output.
The loss function when training the EN is the MSE loss between the actual and
predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN's predicted output, and target output, to the CPN.
See Fig. \ref{fig:training}.

@@

\subsubsection{EN training}
For the EN to be useful, it must accurately predict the effect of stimulation
produced by the CPN. In addition to that, backpropagating through it must yield
gradients which train the CPN to output more useful stimulus. We discovered that
this latter property does not necessarily occur simply by virtue of the former.
We have found that an EN can be trained to extreme levels of predictive power, even
on white noise stimulation - to orders of magnitude lower loss than the task loss -
while at the same time backpropagation through it yields gradients which are not
useful for training the CPN. As a result, the CPN exhibits unstable training.
We have thus far been unable to discover a way to measure and optimize for this second
property, though pursuing it may be a fruitful subject of future ML research. Instead
of attempting to train an EN directly to exhibit this property, we discovered a
training regime that reliably makes it appear.

The central concept of our training regime is to provide a carefully chosen variety of
training examples, and to use heavy weight regularization, in order to force
the EN to generalize well to stimulation which it hasn't yet seen. We hypothesize that
the crux of our EN training problem is one of over-fitting: that the EN may be trained
to great predictive power on a set of stimulation examples, but that the effective
dimensionality of the problem it is learning is comparatively quite high, due to both
the dimensionality of our stimulation parameters $\theta$, and the dynamics of the
network being stimulated. As a result, the hypersurface which it has fit to the training
examples may exhibit gradients not reflective of reality.

First, and most significantly, we structure the training data set for each EN training epoch
in a very specific way. Each epoch, we include examples from the current CPN, the current CPN
with Gaussian-distributed mean-0 noise added to its parameters, and examples of white noise
stimulation. We developed this structure under the believe that we needed to cover a sufficient
variety of examples to prevent overfit, and to do so in a way that emphasizes the neighborhood
in CPN parameter space of the current CPN. We initially attempted to pad our dataset with white
noise examples alone, but that was not sufficient to stabilize CPN training, despite the EN's
ability to reach a high predictive power. Adding in the addition of samples from the parameter
space near the current CPN appears to help ensure that the EN is localized enough to give
useful gradients for that area. As we will explain below, this produces an additional problem
that the EN is to specialized to the local area of CPN parameter space, but we then solve
that problem using EN retraining, interleaved with the CPN, as we will explain further below.

Thus, we composed each batch of training data in the following way:

\begin{table}[]
\begin{tabular}{|l|l|}
\hline
\textbf{Source}                                                                                                                            & \textbf{Proportion of dataset} \\ \hline
Examples produced by the current CPN                                                                                                       & 10\%                           \\ \hline
\begin{tabular}[c]{@{}l@{}}Examples produced by a copy of the current CPN,\\ with mean-0 random noise added to its parameters\end{tabular} & 60\%                           \\ \hline
Examples of white noise stimulation                                                                                                        & 30\%                           \\ \hline
\end{tabular}
\end{table}

Second, we use of PyTorch's $AdamW$ optimizer, which includes an option for weight regularization,
and a carefully chosen learning rate schedule. Our intuition is to leverage regularization as
one of the standard mitigations for overfit. To converge, the learning rate schedule also needs
to be chosen carefully: a relatively high learning rate at first, dropping to a very low rate soon
thereafter. Using too-low or too-high of a learning rate in any phase causes the EN learning
to not converge. EN training proceeds until it reaches a threshold of prediction error, defined as
a fraction of the current CPN's task loss.

\subsubsection{CPN training}
Once we discovered an effective way to train an EN, CPN training was straightforward. For this phase,
We generate training examples using the current CPN alone. The EN generates predictions of the
effect of those stimulations, and we backpropagate through the EN to generate training gradients
for the CPN. As with the EN, we constitute the dataset with a large random sample of trials from
the original Michaels task. See Fig. \ref{fig:training}.

As with the EN, the choice of learning rate schedule for training the CPN is somewhat important.
The CPN appears to train in two phases. In the first phase, it is largely learning the structure
of the reach-to-grasp task, e.g. that the muscles need to stay still until the reach begins. During
that phase, the learning rate can be quite high. Later, the CPN begins to learn the mapping between
the object shape information observed from the simulated brain, and how that should map onto
stimulation. That phase of the training takes much longer, and requires a learning rate 2-3
orders of magnitude lower than in the first phase. We will cover this in some more detail in
Section \ref{sec:results}.

\subsubsection{Interleaved CPN/EN training, and adapting to non-stationarity}
Having defined the training epochs for the EN and CPN, we can define a training
algorithm on the basis of those two. We train the two in alteration, creating
a new EN each time we enter an EN training period. We explored the possibility
of reusing an existing EN by retraining it, for the purpose of training efficiency.
However, we found that retraining an EN was no more fast than training a new one,
and in-fact usually took longer, despite significant experimentation with learning
rates.

The key challenge we face is determining when an EN is no longer suitable for training the
current CPN. Expiring an EN at the right time accomplishes two things. First, it ensures
that our current EN is trained properly for the current CPN; without that, the CPN training
will become unstable. Second, expiring the EN is how we adapt our learning to the brain's
non-stationarity. At some point, the brain will have changed sufficiently for our EN to be
out-of-date, and therefore requiring replacement.

We solve this challenge through a simple set of metrics, which, when satisfied, indicate
we need to transition to an EN training epoch. Our first metric simply captures the EN's
declining predictive power. When the EN's prediction error reaches a point which is
sufficiently above some ratio to the CPN's task loss, we expire the EN immediately.
Second, we expire the EN if CPN loss does not improve (or gets worse) across a sufficient
number of recent training steps. This resembles common stopping conditions in
iterative learning: one stops when learning is no longer improving the model. Together,
these two metrics appear effective in ensuring we expire the EN at the right time.

Unfortunately, because we expire the EN rather than retraining it, and because we need
to expire it so often, training is less efficient than it otherwise could be. We explore
this some in Section \ref{sec:results}. In the next subsection, we illustrate one way we
mitigate training inefficiency.

TODO: pseudocode

\subsubsection{Optimization: data reuse}
One obvious

\section{Results}
\label{sec:results}

@@

Using the simulation described above, we perform three experiments to
demonstrate the co-processor's ability to learn:
\begin{itemize}
	\item 50\% AIP loss
	\item 100\% connectivity lesion between the F5 and M1 modules
	\item 100\% connectivity lesion between the F5 and M1 modules without brain co-adpatation
	\item 50\% M1 loss
\end{itemize}

@@
